thesis2024

### 2025-05-02 thesis2024.tex
Macbook session. Tests to latex chapter structure 01

\setstretch{2}
\bookmarksetup{startatroot}

@@

\chapter{Creating frames of reference for chert exploitation during the
Late Pleistocene in Southwesternmost
Iberia}\label{creating-frames-of-reference-for-chert-exploitation-during-the-late-pleistocene-in-southwesternmost-iberia}

@@

\section*{Abstract}\label{abstract-1}
\addcontentsline{toc}{section}{Abstract}

@@

\section{Introduction}\label{introduction-1}

@@

\section{Geological setting and chert
groups}\label{geological-setting-and-chert-groups}

\subsection{Geological setting}\label{geological-setting}

@@

\subsection{Chert outcrops}\label{chert-outcrops}

@@

\section{Materials and methods}\label{materials-and-methods}

@@

\section{Results}\label{results}

@@

\subsection{Western Algarve}\label{western-algarve}

@@

\subsection{Eastern Algarve}\label{eastern-algarve}

@@

\subsection{Other outcrops}\label{other-outcrops}

@@

\section{Discussion}\label{discussion}

@@

\section{Conclusion}\label{conclusion}

@@

\textbf{Authors}: Joana Belmiro, Xavier Terradas, Salvador
Dominguez-Bella, João Cascalheira

@@

\chapter{From Stone to Tool}\label{from-stone-to-tool}

>>>>

\setstretch{1.5}
\bookmarksetup{startatroot}

@@

\chapter*{Abstract}\label{abstract-1}
\addcontentsline{toc}{chapter}{Abstract}

\markboth{Abstract}{Abstract}

\chapter{Creating frames of reference for chert exploitation during the Late Pleistocene in Southwesternmost Iberia}

@@

@@

\section*{Introduction}\label{introduction-1}
\addcontentsline{toc}{section}{Introduction}

\markright{Introduction}

@@

\section*{Geological setting and chert
groups}\label{geological-setting-and-chert-groups}
\addcontentsline{toc}{section}{Geological setting and chert groups}

\markright{Geological setting and chert groups}

\subsection*{Geological setting}\label{geological-setting}
\addcontentsline{toc}{subsection}{Geological setting}

@@

\subsection*{Chert outcrops}\label{chert-outcrops}
\addcontentsline{toc}{subsection}{Chert outcrops}

@@

\section*{Materials and methods}\label{materials-and-methods}
\addcontentsline{toc}{section}{Materials and methods}

\markright{Materials and methods}

@@

\section*{Results}\label{results}
\addcontentsline{toc}{section}{Results}

\markright{Results}

@@

\subsection*{Western Algarve}\label{western-algarve}
\addcontentsline{toc}{subsection}{Western Algarve}

@@

\subsection*{Eastern Algarve}\label{eastern-algarve}
\addcontentsline{toc}{subsection}{Eastern Algarve}

@@

\subsection*{Other outcrops}\label{other-outcrops}
\addcontentsline{toc}{subsection}{Other outcrops}

@@

\section*{Discussion}\label{discussion}
\addcontentsline{toc}{section}{Discussion}

\markright{Discussion}

@@

\section*{Conclusion}\label{conclusion}
\addcontentsline{toc}{section}{Conclusion}

\markright{Conclusion}

@@

\textbf{Authors}: Joana Belmiro\textsuperscript{1}, Xavier
Terradas\textsuperscript{2}, Salvador
Dominguez-Bella\textsuperscript{3}, João Cascalheira\textsuperscript{1}

\textsuperscript{1} Interdisciplinary Center for Archaeology and the
Evolution of Human Behaviour (ICArEHB), University of Algarve;
\textsuperscript{2} CSIC; \textsuperscript{3} UCA

\textbf{Reference}: xxx

@@

\chapter{From Stone to Tool: How Raw Materials Influenced Upper
Paleolithic Technology in Southwestern Iberia (Vale
Boi)}\label{from-stone-to-tool-how-raw-materials-influenced-upper-paleolithic-technology-in-southwestern-iberia-vale-boi}

\textbf{Authors}: Joana Belmiro, Nuno Bicho, Xavier Terradas, João
Cascalheira

\section{Abstract}\label{abstract-3}

The Upper Palaeolithic (UP) of westernmost Europe was marked by
technological and cultural transformations and abrupt climatic shifts.
The Iberian Peninsula, particularly southwestern Iberia, served as a
refugia and key eco-cultural niche, making it a key region for UP
studies. Vale Boi, the only site in southwestern Iberia with a nearly
complete UP sequence, provides critical insights into technological,
economical and mobility patterns over time. This study examines lithic
raw material procurement and technological organisation at Vale Boi,
focusing on the differences between local and non-local cherts. We
analysed chert assemblages from Gravettian, Proto-Solutrean, and
Solutrean levels, integrating previously published techno-typological
and raw material macroscopic and petrographic data. Our results indicate
that non-local cherts were intensively reduced. The presence of varied
techno-typological classes suggests that non-local cherts were not
exclusively transported as finished tools, instead highlighting their
role as versatile, transportable raw material volumes. Gravettian
occupations exhibited greater reliance on non-local cherts, suggesting
short-term occupations and higher mobility, while Proto-Solutrean and
Solutrean occupations showed increased dependence on local materials,
suggesting long-term occupations. These findings expand our
understanding of UP technological adaptations in southwestern Iberia,
emphasizing the complex role of lithic resources in mobility,
settlement, and social networks.

\textbf{Keywords:} Upper Palaeolithic; Lithic technology;
Hunter-gatherer mobility; Chert procurement; Iberian Peninsula

\section{Introduction}\label{introduction-3}

The Upper Palaeolithic (UP) in Europe, dating to approximately
40,000--10,000 years ago, was marked by significant technological
innovations, shifts in subsistence strategies, and the emergence of
diverse regional traditions (Gamble et al. 2004; Straus 1995).

In westernmost Europe, five main technocomplexes have been identified
within the UP - Aurignacian, Gravettian, Proto-Solutrean, Solutrean, and
Magdalenian - each contributing new evidence on resilient behaviours and
cultural continuity. These technocomplexes are distinguished by regional
adaptations as well as technological and subsistence innovations
(\textbf{baker2024?}; Nuno Bicho et al. 2017a; \textbf{bicho2011?};
\textbf{bicho2008?}; \textbf{cascalheira2017?}; Straus 1991;
\textbf{straus1993?}; \textbf{zilhão1997?}).

These innovations are believed to have resulted from human adaptations
and resilience to the shifting conditions of the Late Pleistocene, which
was marked by abrupt climatic events such as the Heinrich events and the
Last Glacial Maximum (LGM) (Sanchez Goñi and Harrison 2010b). These
climatic shifts had a profound impact on flora and fauna, with exception
of spatially limited refugia, likely disrupting settlement patterns and
the organisation of hunter-gatherer groups, ultimately leading to a
process of cultural reorganisation (\textbf{bradtmöller2012?}; Morin
2008; \textbf{schmidt2012?}), which has frequently been a focus of UP
research.

In particular, the Iberian Peninsula has been recognized as a key region
for studying UP adaptations and cultural transitions (e.g.,
\textbf{aubry2016?}; \textbf{aubry2022?}; \textbf{bicho2011?};
\textbf{bicho2017a?}; \textbf{bradtmöller2012?};
\textbf{cascalheira2017?}; \textbf{cascalheira2021?};
\textbf{cascalheira2013?}; \textbf{finlayson2006?}; Gamble et al. 2004;
\textbf{jennings2011?}; \textbf{jiménez-espejo2007?};
\textbf{linscott2023?}; \textbf{rasilla2020?}; \textbf{schmidt2012?};
\textbf{zilhão1997?}; \textbf{zilhão2021?}; \textbf{zilhão2017?}). This
significance is largely due to its location in southwestern Europe,
which enabled it to maintain a relatively temperate climate even during
the coldest periods, such as the LGM. As a result, the Iberian Peninsula
served as one of Europe's most important glacial refugia
(\textbf{gómez2007?}; \textbf{gonzález-sampériz2010?};
\textbf{hewitt2000?}; \textbf{jennings2011?}; \textbf{schmidt2012?}).

These refugia provided stable environments that were continuously
occupied by hunter-gatherers, leading to the formation of eco-cultural
niches - stable socio-ecological systems where human settlements were
recurrent, and cultural adaptations were shaped by ecological
constraints and opportunities (\textbf{cascalheira2017?}). These
eco-cultural niches present a unique opportunity to examine long-term
dynamics in resource exploitation, offering valuable insights into human
adaptive strategies in response to cultural and social transformations.

More specifically, southwesternmost Iberia has been identified as a key
eco-cultural niche for studying cultural transitions and the role of
this peripheral coastal region within hunter-gatherer social networks.
This is mainly due to the archaeological site of Vale Boi, the only site
in southwestern Iberia with a nearly complete UP sequence, making it a
crucial reference for examining long-term technological and mobility
patterns in the region (\textbf{cascalheira2017?}). Its well-preserved
stratigraphic continuity allows for a detailed analysis of cultural and
technological trends across different phases of the UP.

Building on this significance, research at Vale Boi has provided
valuable insights into hunter-gatherer subsistence strategies and
technological organisation (\textbf{belmiro2021?}; \textbf{bicho2013?};
\textbf{bicho2017a?}; \textbf{cascalheira2019?}; \textbf{horta2019?};
\textbf{manne2012?}; \textbf{marreiros2015?}; \textbf{pereira2016?}),
and lithic technology studies from this site have highlighted the
region's shifting and complex role within the broader territorial
landscape.

In this context, comparisons between Vale Boi and the Gravettian
occupations in central Portugal and southern Spain revealed marked
regional differences, suggesting more limited movement of people and
information in southwestern Iberia. These differences were suggested to
be related to Vale Boi's peripheral location and the distinct ecological
characteristics of westernmost Algarve (\textbf{marreiros2013?}). In
contrast, Proto-Solutrean evidence from the site indicates technological
similarities and strong connections with central Portugal
(\textbf{belmiro2021?}), while the Solutrean period stands out as a
highly dynamic phase at Vale Boi, during which the site functioned as a
key connection point between central Portugal and southern Spain,
facilitating social networks and cultural exchanges
(\textbf{bicho2017a?}; \textbf{cascalheira2013c?}).

Expanding on these findings, recent research has further refined our
understanding of mobility and social networks through lithic raw
material studies at the site. Our team's recent work identified
non-local cherts originating from central Portugal and southern Spain in
the UP lithic assemblages of Vale Boi (\textbf{belmiro2025?}). The
proportions of these materials varied throughout the UP with earlier
occupations (Gravettian) showing high frequencies of non-local chert,
while later phases (Proto-Solutrean and Solutrean) exhibited a marked
decline. These findings suggest the establishment and maintenance of
extensive social networks connecting southwestern Portugal to southern
Spain and central Portugal as early as the beginning of the UP.

These results, particularly those related to the Gravettian, were
interpreted as outcomes of residential mobility rather than direct
social network influences, though further investigation is required.

While previous studies have explored Vale Boi's technological
organisation, a direct analysis of how raw material sourcing shaped
lithic production strategies remains missing. This study bridges that
gap by examining chert procurement in relation to tool production, use,
and discard, providing critical insights into hunter-gatherer mobility,
settlement strategies, and social interactions across Iberia.

This paper presents a combined analysis of lithic technology and raw
material sourcing, aiming to refine our understanding of technological
and social adaptations throughout the UP in southwestern Iberia. By
integrating archaeological and ethnographic perspectives, we test
expectations regarding the organisation of hunter-gatherer technological
systems and their implications for mobility, resource management, and
social interactions. In doing so, we contribute to ongoing discussions
about eco-cultural niches and the role of southwestern Iberia within
broader UP social networks.

\subsection{Theoretical framework and
objectives}\label{theoretical-framework-and-objectives}

Lithic technological organisation has long been a central focus in
archaeology, offering critical insights into past human behaviour and
adaptive strategies. It encompasses all stages of stone tool production
and use, from raw material acquisition to discard. Nelson
(\textbf{nelson1991?}) defines it as the selection and integration of
strategies for tool production, use, transport, and maintenance, while
Kelly (\textbf{kelly1988?}) emphasizes its spatial and temporal
dimensions within cultural systems. These perspectives highlight the
interrelationship between raw material sourcing, mobility, and
settlement organisation---key factors in understanding hunter-gatherer
adaptations.

Raw material procurement is a fundamental aspect of lithic technological
organisation, as it played a crucial role in the lives of past
hunter-gatherer groups by determining the possibilities for tool
production and use (\textbf{bamforth1997?}). Approaches focused on the
study of raw materials provide key knowledge on several topics of
hunter-gatherer organisation such as procurement and mobility strategies
(\textbf{ambrose1990?}; \textbf{binford1979?}; \textbf{binford1985?};
\textbf{gould1985?}; \textbf{kuhn1991?}; \textbf{mccall2007?};
\textbf{gramly1980?}), occupation types and their duration
(\textbf{torrence1983?}; \textbf{grove2023?}; \textbf{kuhn2004?};
\textbf{surovell2009?}), and social networks and exchanges
(\textbf{gamble1999?}; \textbf{whallon2006?}). However, raw material
studies often focus on the characterisation of geological and
archaeological raw materials to identify local and non-local raw
materials within an assemblage, their possible sources, and examine
their proportions within a site over different occupations and periods.

While distance to source and raw material proportions offer valuable
insights, they represent only one aspect of raw material procurement and
management. A raw material-focused approach alone is insufficient for
comprehensively understanding mobility patterns and technological
organisation. Numerous studies and models from both ethnographic and
archaeological perspectives have demonstrated the complex interplay
between lithic technological variability, raw material provisioning, and
mobility.

For example, technological variability within a lithic assemblage may be
explained by the abundance and quality of raw materials
(\textbf{andrefsky1994?}; \textbf{brown1999?}; \textbf{oestmo2017?}).
Whenever raw materials in proximity to a residential site are abundant
and of good quality, a high percentage of formal tools are produced
(\textbf{andrefsky1994?}; \textbf{perlès1992?}), and artefacts show less
reduction intensity and decreased conservation through retouching
(\textbf{thacker2002?}). Whenever local raw materials are scarce or of
low quality, non-local resources are procured to produce mostly formal
tools (\textbf{andrefsky1994?}; \textbf{perlès1992?}). Similarly, when
groups move to areas where raw material is scarcer or of poorer quality,
more effort is placed into the preparation of transportable cores and
the production of tools (\textbf{kelly1988?}).

However, the different patterns between local and non-local raw material
proportion and use in lithic assemblages may be less obvious whenever
prehistoric groups have easy access to non-local raw materials due to
high mobility or exchanges between groups (\textbf{akerman2002?};
\textbf{contexts1982?}; \textbf{torrence1986?}; \textbf{brown1999?};
\textbf{oestmo2017?}). Furthermore, Roth and Dibble (\textbf{roth1998?})
demonstrate that even in areas where good quality raw material is
abundant, non-local raw materials were still transported to the site,
often in the form of large blanks or trimmed cores, and both local and
non-local raw materials were retouched into tools.

Tools, flakes and cores often functioned as portable sources of raw
material within mobile toolkits or during non-local raw material
procurement, enabling stone knapping during periods of limited time or
in environments with scarce or unsuitable materials
(\textbf{gould1971?}; \textbf{kelly1988?}; \textbf{kuhn1994?};
\textbf{morrow1996?}; \textbf{perlès1992?}; \textbf{shott1986?}). In
this context, the transport of formal cores (e.g., blade cores) and
bifaces used as cores have been interpreted as lightweight and
reconfigurable raw material storage strategies (\textbf{clark1987?};
\textbf{kelly1988?}).

Formal tools and the degree of tool maintenance have also been
associated with planned preparation, efficient use, and
transportability, reflecting their connection to mobile settlement
strategies, short-term site occupations (\textbf{torrence1983?}) and the
regional distribution of raw materials (\textbf{bamforth1986?}). In some
cases, non-local raw materials, transported as part of personal toolkits
(frequently associated with residential mobility), would primarily
arrive at sites as retouched pieces (\textbf{kuhn2004?}). These
artefacts would show higher degrees of reduction and reworking
(\textbf{kuhn2004?}) and a higher percentage of non-cortical flakes
(\textbf{roth1998?}), as they would have been in use for an extended
period before being discarded. When knapped on-site, non-local materials
also tended to exhibit higher proportions of retouched tools compared to
local materials.

Tool variability and size are also impacted by mobility and uncertainty
regarding raw material availability: increased group mobility may create
a pattern of limited tool variability, and tools may become less
specialised and easily reconfigurable, such as retouched blanks
(\textbf{odell1981?}; \textbf{shott1986?}; \textbf{siegel1984?}).
Increased mobility also often results in smaller tools
(\textbf{shott1986?}), although other authors have noted that large
tools may be well-suited to a mobile lifestyle due to their extended
use-life, allowing them to remain functional in areas where raw
materials are scarce or unsuitable for knapping (\textbf{morrow1996?}).

The proportions of local vs non-local raw materials discussed thus far
have focused greatly on group mobility, which is linked with residential
mobility and occupation duration. Residential mobility (e.g., foraging
range size and frequency of residential moves) is a defining
characteristic of hunter-gatherer societies and plays a central role in
their lifeways (\textbf{kelly1995?}). It is a key factor driving changes
in lithic assemblages (\textbf{brown1999?}; \textbf{oestmo2017?};
\textbf{roth1998?}) as it significantly impacts the cost and use of
artefacts and raw materials since they affect the relation between the
distance of raw material sources to sites (\textbf{kuhn2004?}). As such,
different raw material and technological patterns have been observed
between short-term and long-term occupations.

Short-term occupations typically rely on transported (non-local)
toolkits (\textbf{torrence1983?}) and raw materials
(\textbf{surovell2009?}), show higher reduction intensity, and an
increased tools-to-debitage ratio in the assemblages
(\textbf{grove2023?}). Long-term occupations typically rely on
provisioning sites with locally available and suitable raw materials,
while non-local materials are gradually used and discarded
(\textbf{grove2023?}; \textbf{surovell2009?}). Non-local raw material
cores are highly reduced and extensively exploited (\textbf{kuhn2004?};
\textbf{roth1998?}) and exhibit higher blank-to-core ratios
(\textbf{roth1998?}).

Altogether, different studies highlight significant variability in
lithic assemblages and raw material proportions, reflecting different
facets of hunter-gatherer technological organisation. This variability
appears between behaviourally and ethnically similar sites
(\textbf{bamforth1986?}), across distinct chrono-cultural periods, and
even within a single site among various raw materials. To fully
understand past hunter-gatherer behaviours, organisation, and lifeways,
it is crucial to examine all stages of lithic production--- procurement,
distribution, use, maintenance, and discard---through interconnected
approaches that combine lithic technology and raw material analyses.

This study aims to refine our understanding of lithic resource
provisioning, use, and management strategies among the UP
hunter-gatherer groups settled at Vale Boi. To examine the role of chert
procurement and reduction strategies at Vale Boi we test several
expectations derived from the previously mentioned studies and models
about raw material use and lithic technology organisation, focusing
mainly on the different management of local and non-local raw materials,
associated with mobility and occupation length, which are summarised in
Table~\ref{tbl-expectations}. These expectations are tested through the
comparison of local and non-local chert technological and metric
characteristics.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.3986}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.6014}}@{}}

\caption{\label{tbl-expectations}Expected results of the technological
and metric comparison between local and non-local cherts.}

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
1. Local raw materials
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
2. Non-local raw materials
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
1a) Low use-lives, with knapping strategies characterised by
manufacture, use and discard & 2a) Majority of formal tools, in the case
of individual provisioning or higher use-lives, with knapping strategies
characterised by the manufacture, use and maintenance \\
1b) Decreased reduction, larger cores and debitage, lower blank-to-core
ratio & 2b) Increased reduction, smaller cores and debitage, higher
blank-to-core ratio \\
1c) Decreased tool reworking, low presence of retouch and/or tool
maintenance, and lower tool-to-debitage ratio & 2c) Increased tool
reworking, high presence of retouch and/or tool maintenance, and higher
tool-to-debitage ratio \\
1d) Higher tool typology diversity & 2d) Lower tool diversity,
characterised by less specialised tools \\

\section{Materials and methods}\label{materials-and-methods-2}

\subsection{Site description}\label{site-description-1}

Vale Boi is located on the western coast of the Algarve region, southern
Portugal, at a latitude of N 37º5'24.1'' and longitude of W 8º48'32.4''
(Gibaja and Bicho 2006, the original site coordinates were converted
from Degree/Minute to Degree/Minute/Second), within a small valley
extending \textasciitilde2 km southward to the Atlantic Ocean
(Figure~\ref{fig-map} a). The site is nestled among limestone outcrops
that create west- and southwest-facing rock shelters
(Figure~\ref{fig-map} b). Covering more than 10,000 square meters along
the valley slope, Vale Boi includes three main excavated areas, Slope,
Terrace, and Shelter (Figure~\ref{fig-map} c), which were excavated
between 2000 and 2019 (\textbf{bicho2007?}; Nuno Bicho and Haws 2012;
\textbf{bicho2013?}; \textbf{cascalheira2017?};
\textbf{cascalheira2013?}; \textbf{manne2012?}). Based on the available
raw material and technological evidence, particularly the macroscopic
and petrographic data (\textbf{belmiro2025?}), the Terrace and Shelter
areas were selected for the current study.

\includegraphics{figures/from-stone-to-tool/VB_map.png}

\caption{\label{fig-map}a) Location of the archaeological site of Vale
Boi (orange box) within the Iberian Peninsula; b) Photo of the limestone
hill with terrace and shelter locations; c) Schematic of the site area.
Figures b-c) were adapted from Cascalheira et al., 2017. Figure a) was
produced with resource to the open-source software QGIS
(https://qgis.org/)}

The Terrace area preserves occupations ranging from the UP to the Early
Neolithic. While excavations started in 2000, an additional
8-square-meter trench was opened in 2012 to refine the stratigraphic
sequence and assess the presence of earlier UP deposits. Between 2012
and 2016, eight primary litho-stratigraphic units were identified. The
UP sequence includes multiple occupations, spanning
\textasciitilde10,000 years: Gravettian (levels 8 to 6) dated between
approximately c.~32 and 27 ka cal BP, Proto-Solutrean (levels 5 to 4E)
between approximately c.~26 and 24 ka cal BP, and Solutrean (levels 4D,
4C, 4, and Lower 3) between approximately c.~24 and 20 ka cal BP
(\textbf{belmiro2021?}; \textbf{cascalheira2017?};
\textbf{cascalheira2013b?}).

The Shelter area contains four main lithostratigraphic units, with
occupations from the Magdalenian, Solutrean, and Gravettian periods. The
Magdalenian and Gravettian levels (Z and D, respectively) were partially
eroded, while the Solutrean levels represent the most intensive
occupation, with three well-preserved archaeological horizons (layers A
to C), dated to approximately c.~24-22 ka cal BP. These occupation
levels were found beneath blocks of limestone that likely collapsed from
the rock shelter ceiling, probably shortly after the LGM
(\textbf{cascalheira2010?}; \textbf{cascalheira2017?}).

The site's lithic assemblages represent all stages of stone tool
production, with raw material use varying by occupation. Technological
analyses of these assemblages have provided insights into the adaptive
strategies of UP hunter-gatherers at Vale Boi, revealing technological
shifts across different cultural phases. Given the chosen sample for
this study (see Section Datasets and analysis), we summarise here the
results for the technological analysis of the Solutrean assemblages
(levels A-C) of the Shelter (\textbf{cascalheira2010?};
\textbf{cascalheira2013a?}) and the Proto-Solutrean assemblages (levels
4E-5) of the Terrace (\textbf{belmiro2020?}; \textbf{belmiro2021?}). The
Gravettian occupations (levels 6-8) of the Terrace have been recently
studied, but the results remain unpublished.

The lithic analysis of the Solutrean levels from the Shelter (levels
A-C) shows a high number of lithic artifacts, with over 22,000 artifacts
analysed (including chips which correspond to \textasciitilde60\% of the
assemblage and shatter \textasciitilde17-19\%). The volumes introduced
into the site were mostly cortical, given the high frequencies of cortex
in blanks and retouched tools. The knapping strategies are dominated by
unidirectional and bidirectional prismatic cores, with frequent core
maintenance products. Flakes dominate the debitage assemblage, and
endscrapers, notches, splintered pieces and retouched blanks are the
most common retouched tools. Stemmed and winged points, laurel leaves,
and shouldered points were also identified (\textbf{cascalheira2013a?}).

The Proto-Solutrean horizons yielded more than 26,500 artifacts
(including chips which correspond to \textasciitilde65-70\% of the
assemblage and shatter \textasciitilde20\%), \textasciitilde11,000
attributed to the bottom of Terrace's level 5, and \textasciitilde15,600
attributed to the top of level 5 and level 4E. Both horizons are
characterised by knapping strategies dominated by unidirectional
prismatic cores, with little platform preparation, focused on the
obtention of bladelets (in bottom level 5) and flakes. Retouched tools
are dominated by splintered pieces, endscrapers and retouched blanks,
with Vale Comprido points (the Proto-Solutrean fossil director) being
barely inexistent in chert and mostly produced in dolerite
(\textbf{belmiro2021?}).

Previous studies showed that the main raw materials used at the site are
chert, quartz, greywacke and smaller amounts of chalcedony, with the
percentages of these changing across occupations. Notably, the lower
Proto-Solutrean levels show high frequencies of quartz
(\textbf{belmiro2021?}), whereas chert becomes the dominant raw material
in the Solutrean occupations of the Shelter (\textbf{cascalheira2010?}).
Likewise, different raw materials appear to have been used for specific
purposes, with chert being particularly associated with the production
of high percentage of blanks and retouched tools, and greywacke being
extensively used for anvils (\textbf{horta2019?}).

Quartz and greywacke are locally available from streambeds near the
site, whereas regional chert and chalcedony sources lie further afield,
within a \textasciitilde20 km radius (\textbf{pereira2016?}). Our
previous studies showed, as expected, the existence of local cherts in
the assemblages (Figure~\ref{fig-cherts} a), originating from the Lower
Jurassic (cherts) and Triassic (chalcedony) outcrops in proximity to the
site (\textbf{belmiro2025?}). Despite petrographic differences between
cherts and chalcedony, for this study any chert varieties (e.g.,
chalcedony or jasper) are referred to as chert (\textbf{luedtke1978?}).
A significant portion of non-local cherts was also identified
(Figure~\ref{fig-cherts} b-c), with some identified sources, such as the
Cretaceous outcrops of central Portugal (Lisbon and Rio Maior areas) and
the Upper Jurassic oolitic/peloidal cherts of the Betic Systems in
southern Spain. This indicates varying degrees of contact or long-range
mobility throughout the UP, likely influenced by shifts in settlement
patterns and the duration of site occupation.

\includegraphics[width=5.20833in,height=\textheight]{figures/from-stone-to-tool/types_short.png}

\caption{\label{fig-cherts}Simplified scheme of chert types identified
in the UP sequence of the Terrace and Shelter areas of Vale Boi: a)
local chert macroscopic variability (T1 to T5), including chalcedony
(T1; top left); b) non-local chert types present in all occupations
except for T11 which is only present in the Solutrean; c) TL09 (left)
and TL06 (right) varieties within the trace lithologies group (TL)
composed of non-local cherts with 1-3 samples per TL type and limited to
specific archaeological levels and areas. Summarised from Belmiro et
al., (2025)}

Previous research suggests that both the Terrace and Shelter functioned
as seasonal residential camps, repeatedly occupied for extended periods.
This interpretation is based on high-density lithic assemblages,
evidence of tool manufacture, heat-cracked rocks associated with grease
rendering, diverse faunal remains (marine and terrestrial), and personal
ornaments and portable art (\textbf{manne2012?}). This is true for most
of the UP occupations, although with some exceptions.

Cascalheira et al.~(2017) applied the Whole Assemblage Behavioural Index
(WABI) (\textbf{riel-salvatore2004?}; \textbf{riel-salvatore2010?}) to
reconstruct and compare mobility strategies at Vale Boi. By analysing
lithic volumetric density and retouched tool frequencies from the
Shelter's Solutrean (\textbf{cascalheira2010?};
\textbf{cascalheira2013a?}) and Magdalenian (\textbf{mendonça2009?})
levels, as well as the Terrace's Gravettian levels
(\textbf{marreiros2015?}), the authors confirmed that most occupations
corresponded to long-term residential base camps, composed of
high-density assemblages with low degree of retouch, except for the
Early Gravettian and Magdalenian occupations. Regarding the Terrace
area, however, the results were based on previous excavations
(\textless2012), and the identification of new layers and reorganisation
of the occupations is yet to be tested through the use of lithic
techno-typological methods.

A similar conclusion was suggested based on the interpretation of the
chert sourcing and proportions in the several UP occupations of the
Terrace and Shelter of Vale Boi and following other models relating
local/non-local raw materials with mobility (\textbf{grove2023?};
\textbf{surovell2009?}). The high proportions of non-local chert
(\textasciitilde50\%) in the Gravettian occupations (levels 6 and 7),
and gradual reduction in the following occupations, were interpreted as
a possible result of settlement changes, where the Gravettian
occupations corresponded to short-term occupations, and the other
occupations corresponded to residential, long-term occupations with more
infrequent moves (\textbf{belmiro2025?}).

\subsection{Datasets and analysis}\label{datasets-and-analysis}

The current study focuses on the correlation between the
techno-typological analysis and the geoarchaeological study of the chert
lithic assemblages of the UP of Vale Boi. It integrates previously
collected datasets, which have been published or are openly accessible
in online repositories, allowing for transparency and reproducibility.

The macroscopic and petrographic data of cherts was previously collected
and published by our team (\textbf{belmiro2025?}) and can also be found
in an online repository
(\url{https://www.doi.org/10.17605/OSF.IO/DBFT2}). This study focused on
the UP chert assemblages of Vale Boi, especially the Gravettian (levels
6 and 7), Proto-Solutrean (levels 4E and 5) and Solutrean (levels 4 to
4D) from the Terrace area and the Solutrean (levels A to C) from the
Shelter area. The methodology used in the study, dataset and supporting
documents can be found in the article and the online repository.

The Solutrean assemblage technological analysis was published by
Cascalheira (\textbf{cascalheira2009?}) as part of a Master's thesis and
following publications (e.g., \textbf{cascalheira2010?};
\textbf{cascalheira2013a?}), with the study of levels A to C of the
Shelter area. The dataset used for this study was not originally
published but can be consulted in our online repository
(\url{http://www.doi.org/10.17605/OSF.IO/UWMPK}).

The Solutrean analysis in this study is limited to the Shelter area due
to the inexistence of technological studies for the better preserved
Solutrean levels of the Terrace (e.g., level 4). However, and as seen in
(\textbf{belmiro2025?}), the raw material patterns seem to show little
differences between the Solutrean in both areas; coupled with the
similar radiocarbon dates between the areas (c.~24-22 and 24-20 ka cal
BP), we assume that they are representative of similar, long-term,
residential occupations which aptly characterise the Solutrean groups of
Vale Boi.

The Proto-Solutrean technological analysis was the result of a Master's
thesis (\textbf{belmiro2020?}) and following publication
(\textbf{belmiro2021?}), with the study of levels 4E and 5 of the
Terrace area. The dataset and analysis can be consulted in the original
online repository (\url{http://www.doi.org/10.17605/OSF.IO/456EG}).

Finally, the Gravettian technological analysis was the result of a
recent and unpublished study, focusing on the chert lithic artefacts
from levels 6 and 7 of the Terrace area (\textgreater{} 2012
excavations). This dataset and preliminary technological analysis
results can be found in our online repository
(\url{http://www.doi.org/10.17605/OSF.IO/UWMPK}) and in Online Resource
1.

One challenge of integrating datasets collected by different researchers
is variability in methodologies and recorded variables. However, this
study minimizes such inconsistencies by ensuring methodological
alignment across datasets, as all data collection followed standardised
attribute-based lithic analysis protocols. The lithic analysis datasets
used in this study adopted an attribute-based methodology, which aims to
describe morphological and metrical attributes of technological classes.
The attributes analysed followed those present in specialised
literature, such as Brézillon (\textbf{brézillon1968?}), and
(\textbf{tixier1980?}), paired with other Palaeolithic lithic attribute
analyses (e.g., \textbf{bicho1992?}; \textbf{zilhão1997?};
\textbf{cascalheira2019?}; \textbf{tostevin2012?}).

After data collection, the databases were imported into R Programming
Environment, where the information was processed through the creation of
descriptive statistical analysis and writing of this article. Some
differences in the collection process, especially between the study by
Cascalheira (\textbf{cascalheira2010?}) and the following studies,
required homogenisation. For example, where in Cascalheira (2010)
retouch was an independent variable (e.g., Class as flake, Retouch as
Yes), in Belmiro et al. (\textbf{belmiro2021?}), retouch was identified
as Class variable. In this situation, the Solutrean dataset was
modified, so that all artifacts with retouch were identified in the
Class variable, instead of the original classification. All changes
applied to the original data made for homogenisation between datasets
can be consulted in the R Studio scripts in our online repository
(\url{http://www.doi.org/10.17605/OSF.IO/UWMPK}).

To integrate technological and raw material data, we merged multiple
datasets using the artifact identifier (``ID''), which corresponds to
total station plot records. This identifier links artifacts to either
precise excavation coordinates (for individually plotted pieces) or
generalised spatial data (for sieved assemblages).

Given that the raw material analysis focused exclusively on individually
piece-plotted artifacts (dimensions \textgreater2 cm), the final dataset
does not include all excavated material but retains the most
diagnostically relevant pieces for analysis. As a result, the final
dataset used in this study does not directly correspond to the full
datasets analysed in the original studies (Table~\ref{tbl-samples}).
Similarly, the raw material study used only currently accessible
artifacts from the assemblages. This is especially relevant for the
Shelter Solutrean chert assemblage, where shatter is absent from the
sample due to the lack of availability for study.

A limitation of this study is its reliance on a subset of the total
lithic assemblage, which excludes small debris. While this may omit some
aspects of on-site knapping activities, the analysed sample remains
representative of the broader technological trends at Vale Boi
throughout the UP sequence.

\begin{longtable}[]{@{}lrr@{}}

\caption{\label{tbl-samples}Sample absolute number comparison between
original technological analysis datasets (chert and chalcedony only,
excluding chips) for the Solutrean occupation of the Shelter area
(levels A-C) published in Cascalheira (2010), the Proto-Solutrean
occupation of the Terrace (levels 4E and 5) published in Belmiro et.
al., (2021) and the Gravettian occupation of the Terrace (levels 6-7)
and the final datasets used in the study after filtering individually
plotted IDs (excluding bucket points) and merging with the
geoarchaeological dataset published in Belmiro et al., (2025).}

\toprule\noalign{}
Assemblage & Original dataset (n) & Merged dataset (n) \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Shelter level A-C & 3290 & 910 \\
Terrace level 4E-5 & 1413 & 635 \\
Terrace level 6-7 & 781 & 766 \\

Following data homogenization and integration, we conducted descriptive
statistical analyses and non-parametric statistical analyses to examine
correlations between raw material use and technological patterns and
test expectations related to the technological organisation of local and
non-local cherts (see Table~\ref{tbl-expectations}), and between
different UP occupations. To test these expectations, we evaluated 9
parameters for each dataset and assemblage; the parameters and their
calculation/measurement methods are detailed in
Table~\ref{tbl-parameters}.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4194}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5161}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.0645}}@{}}

\caption{\label{tbl-parameters}Analysis parameters used in the study,
calculation/measurement used for each parameter, and unit of the
parameter results after analysis.}
